NumPy---Array Computation for Python

### 2020-01-09 body.tex
Fine-tuning outline

The first incarnation of NumPy was a package called Numeric, written in 1995
by Jim Hugunin, then completing his master's thesis at MIT.
Hugunin based his package on previous work by Jim Fulton,
then working at the US Geological Survey, and acknowledges
many others in his initial report of the package \cite{Hugunin-whitepaper}.
Later, Paul Dubois at Lawrence Livermore National Laboratory toke over the project
and funded the writing of the first
manual\footnote{https://conference.scipy.org/scipy2010/slides/travis\_oliphant\_keynote.pdf}.
Numeric provided an array object in Python, written in C, and linking to
standard fast implementations of linear algebra.
% https://stackoverflow.com/questions/26948776/where-did-the-term-broadcasting-come-from/26950256
% https://mail.python.org/pipermail/matrix-sig/1995-November/000143.html
% May want to mention Yorick here (origin of broadcasting idea; it is mentioned later)
% https://aip.scitation.org/doi/pdf/10.1063/1.4823451 
% https://www.dyalog.com/uploads/documents/Papers/Arrays%20of%20Objects.pdf

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{static/sketches/array concepts}
  \caption{\fixme{PLACEHOLDER: This figure needs some text to describe
  it}
   }
  \label{fig:arrayconcepts}
\end{figure*}

@@

Once data is viewed as an $n$-dimensional array, more abstract operations such as
reductions (e.g., summation across columns), products, and so forth, are defined.
Users predominantly interact with NumPy arrays using these operators and
associated utility functions through an easily readable high-level API
(Application Programming Interface).

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{static/sketches/fancy indexing}
  \caption{\fixme{Indexing using arrays, or ``fancy indexing''.}
   }
  \label{fig:broadcasting}
\end{figure}

@@

% NumPy also went from a small community of mostly expert users to
% one of the most widely used foundations of the scientific Python
% ecosystem.
% Now the user base is very different....

\section*{Array protocols}

\begin{figure*}
  \centering
  \includegraphics[width=0.8\textwidth]{static/sketches/duck arrays}
  \caption{\fixme{Examples of so-called ``duck arrays'': arrays
  implemented by another library that behave like NumPy arrays, and
  support NumPy operations.  Through protocols, it is possible to have
  a call such as $\cos(\cdot)$ succeed on each of these.}}
  \label{fig:arrayconcepts}
\end{figure*}

@@

A vast number of projects are built on NumPy;
these projects are consumers of the NumPy API.
Over the last several years, a growing number of projects are providers of
a \emph{NumPy-like API} and array objects targeting audiences with specialized
needs beyond NumPy's capabilities.
For example, the NumPy API is implemented by several popular tensor computation
libraries including CuPy\footnote{\url{https://cupy.chainer.org/}},
JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
and Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}.
PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
and Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
provide tensor APIs with NumPy-inspired semantics.
It is also implemented in packages that support sparse arrays
such as \code{scipy.sparse} and \code{pydata.sparse}.
Another notable example is Dask, a library for parallel computing in
Python.  Dask adopts the NumPy API and therefore presents a familiar
interface to existing NumPy users, while adding powerful abilities to
parallelize and distribute tasks.

>>>>

The first incarnation of NumPy was a package called Numeric, written in 1995
by Jim Hugunin, then completing his master's thesis at MIT.
Hugunin based his package on previous work by Jim Fulton,
then working at the US Geological Survey, and acknowledges
many others in his initial report of the package \cite{Hugunin-whitepaper}.
Later, Paul Dubois at Lawrence Livermore National Laboratory toke over the project
and funded the writing of the first
manual\footnote{https://conference.scipy.org/scipy2010/slides/travis\_oliphant\_keynote.pdf}.
Numeric provided an array object in Python, written in C, and linking to
standard fast implementations of linear algebra.
% https://stackoverflow.com/questions/26948776/where-did-the-term-broadcasting-come-from/26950256
% https://mail.python.org/pipermail/matrix-sig/1995-November/000143.html
% May want to mention Yorick here (origin of broadcasting idea; it is mentioned later)
% https://aip.scitation.org/doi/pdf/10.1063/1.4823451 
% https://aip.scitation.org/doi/pdf/10.1063/1.4822400

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{static/sketches/array-concepts}
  \caption{\fixme{PLACEHOLDER: This figure needs some text to describe
  it}
   }
  \label{fig:arrayconcepts}
\end{figure*}

@@

Once data is viewed as an $n$-dimensional array, more abstract operations such
as reductions (e.g., summation across columns), products, and so forth, are
defined (Figure~\ref{fig:reductions}).
Users predominantly interact with NumPy arrays using these operators and
associated utility functions through an easily readable high-level API
(Application Programming Interface).

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{static/sketches/fancy-indexing}
  \caption{\fixme{Indexing using arrays, or ``fancy indexing''.}
   }
  \label{fig:broadcasting}
\end{figure}

@@

NumPy also went from a small community of mostly expert users to
one of the most widely used foundations of the scientific Python
ecosystem.
Now the user base is very different....
% maybe worth a sentence or two about how we are working on web and
% user docs to make it easier for new users to get started...

\section*{Interoperability and extensibility}

\begin{figure*}
  \centering
  \includegraphics[width=0.8\textwidth]{static/sketches/duck-arrays}
  \caption{\fixme{Examples of so-called ``duck arrays'': arrays
  implemented by another library that behave like NumPy arrays, and
  support NumPy operations.  Through protocols, it is possible to have
  a call such as $\cos(\cdot)$ succeed on each of these.}}
  \label{fig:duck-arrays}
\end{figure*}

@@

A vast number of projects are built on NumPy;
these projects are consumers of the NumPy API.
Over the last several years, a growing number of projects are providers of
a \emph{NumPy-like API} and array objects targeting audiences with specialized
needs beyond NumPy's capabilities (Figure~\ref{fig:duck-arrays}).
For example, the NumPy API is implemented by several popular tensor computation
libraries including CuPy\footnote{\url{https://cupy.chainer.org/}},
JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
and Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}.
PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
and Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
provide tensor APIs with NumPy-inspired semantics.
It is also implemented in packages that support sparse arrays
such as \code{scipy.sparse} and \code{pydata.sparse}.
Another notable example is Dask, a library for parallel computing in
Python.  Dask adopts the NumPy API and therefore presents a familiar
interface to existing NumPy users, while adding powerful abilities to
parallelize and distribute tasks.
